Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-04-18 Manuscript/manuscript.Rmd
manuscript edits

We first evaluated performances for models implemented using OC predictor types, without re-sampling, as these characteristics represented models with the least granularity and could act as a baseline. In evaluating such composite metrics like F1-Score and MCC, we found that these models exhibited poor performances across resolutions, for the GM12878 cell line (Figure 5A). The average F1-Score and MCC across resolution was found to be 0.04 and 0.07 respectively. When including OP perdictor types as the feature space, coupled with no resampling, we found that performances were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). Here, the average F1-Score and MCC was found to be 0.01 and 0.03 respectively. Models built on distance-type predictors in conjuction with no resampling, however, were found to exhibit slightly better performances than the other two types, with average F1-Scores and MCCs of 0.08 and 0.10 respectively (Figure 5C). Importantly, when validating models built on fully imbalanced datasets, the F1-Score and MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, evaluating models using conventional metrics such as AUC or accuracy did not depict the true severity in performance. The average AUC across resolutions for the OC, OP, and distance type predictors was 0.65, 0.66, and 0.85 respectively, where as the average accuracies were 0.89, 0.89, and 0.90 respectively (Supplementary Table 3).

## DATA RE-SAMPLING IMPROVED PERFORMANCE AMONG ALL CLASSES OF MODELS

Comparisons of the effects of the various resampling techniques were then made. When models were built on randomly oversampled data, we found a substantial increase in F1-Score and MCC for both cell lines (Figure 5 D-F). The biggest improvement in performance was seen at the 10 kb resolution, with a sharp downward trend as resolutions decreased. Although performances were low at lower resolutions, both F1-Scores and MCC values were able to be obtained, contrary to models built on completely unbalanced data as seen before. Distance-type predictors were found to offer the highest F1-Scores and MCCs for randomly oversampled data, across all resolutions and for both cell lines.

Further improvement in performance was found when models were implemented on both randomly undersampled (Figure 5 G-H) and SMOTE resampled data (Figure 5 J-K). Although the differences in performance between the two techniques were marginal, SMOTE resampled data was found to yield the best F1-Scores and MCCs among all classes of models develeped and compared in the ensemble framework. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

## DISTANCE-TYPE PREDICTORS PROVIDED THE MOST INFORMATIVE SPATIAL REPRESENTATION OF EPIGENOMIC ANNOTATIONS ASSOCIATED WITH TAD BOUNDARIES

Regardless, of resampling technique however, the inclusion of distance type predictors resulted in better F1-Scores and MCCs when compared to either OC or OP predictor types. This was consistent across all resolutions, for both cell lines. Furthermore, in the best overall model (SMOTE with distance-type predictors) we found that the degradation in model performance as resolution decreased was much more pronounced in the K562 cell line.  on the GM12878 cell line, there was a slight improvement in F1-Score for the 25 kb resolution, followed by a steep decline at 50 kb. Then, both F1-Scores and MCCs showed a momentary increase in performance at the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, the worse being at the 100 kb resolution.

>>>>

We first evaluated performances for models implemented using OC predictor types, without re-sampling, as these characteristics represented models with the least granularity and could act as a baseline. In evaluating such composite metrics like F1-Score and MCC, we found that these models exhibited poor performances across resolutions, for the GM12878 cell line (Figure 5A). <!--The average F1-Score and MCC across resolution was found to be 0.04 and 0.07 respectively.--> When including OP perdictor types as the feature space, coupled with no resampling, we found that performances were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). <!--Here, the average F1-Score and MCC was found to be 0.01 and 0.03 respectively.--> Models built on distance-type predictors in conjuction with no resampling, however, were found to exhibit slightly better performances than the other two types<!--, with average F1-Scores and MCCs of 0.08 and 0.10 respectively--> (Figure 5C). Importantly, when validating models built on fully imbalanced datasets, the F1-Score and MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, evaluating models using conventional metrics such as AUC or accuracy did not depict the true severity in performance (Supplementary Table 3). <!--The average AUC across resolutions for the OC, OP, and distance type predictors was 0.65, 0.66, and 0.85 respectively, where as the average accuracies were 0.89, 0.89, and 0.90 respectively (Supplementary Table 3).--> Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were unable to be reported due to the models inability to produce any true positive cases.

## DATA RE-SAMPLING COUPLED WITH THE INCLUSION OF A DISTANCE-TYPE FEATURE SPACE IMPROVED PERFORMANCES AMONG ALL CLASSES OF MODELS

Comparisons of the effects of the various resampling techniques were then made. When models were built on randomly oversampled (ROS) data, we found a substantial increase in F1-Score and MCC for the GM12878 cell line (Figure 5). The biggest improvement in performance was seen at the 10 kb resolution, with a downward trend as resolutions decreased. Although performances were low at lower resolutions, both F1-Scores and MCC values were able to be obtained, contrary to models built on completely unbalanced data as seen before. Distance-type predictors were found to offer the highest F1-Scores and MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 10). Here, a much sharper negative trend was featured for both F1-Score and MCC and resolution decreased, than in the GM12878 cell line.

Further improvement in performance was found when models were implemented on both randomly undersampled (RUS) and SMOTE resampled data, for all 3 predictor types (Figure 5). Although the differences in performance between the two techniques were marginal, SMOTE resampled data was found to yield the best F1-Scores and MCCs among all classes of models developed and compared in the ensemble framework. As with ROS, distance type predictors yeilded the highest F1-Score and MCC values across resolutions, when applied to either RUS or SMOTE. SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 10). These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do. Likewise, it appears that distance-type predictors provided the most informative spatial representation of epigenomic annotations associated with TAD boundaries.

Moreover, we found that there was a degredation in model performance as data resolution decreased. This was much more pronounced in the K562 cell line. On the GM12878 cell line, there were only marginal differences in performance between 10 kb and 25 kb, followed by a steep decline at 50 kb. Then, both F1-Scores and MCCs showed a momentary increase in performance at 100 kb, the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, starting with highest performance at 10 kb and ending with the worst performance at 100 kb.
